### 2024-03-27 tesis_Ashish/Chapter2/chapter2.tex
update 27 Mar

\chapter{Luminosity measurement at CMS experiment}  %Title of the First Chapter

@@

The CMS experiment employs several devices to measure luminosity, each covering specific pseudorapidity ranges and regions within the CMS detector \cite{pas_18}. The use of several luminometers as shown in Fig.~\ref{fig:lumino_cms} offers a multitude of advantages. Primarily, it ensures redundancy and acts as a backup for potential device failures, ensuring that data collection can proceed without interruption. Additionally, having multiple luminometers allows for cross-checking, providing greater consistency in results and allowing for more reliable data interpretation. These different devices enable better control over systematic uncertainties, as any discrepancies can be identified and adjusted for. Moreover, each device employs different method to measure luminosity, providing complementary information that enriches the overall understanding of particle interactions. Some luminometers recording low rates can be calibrated using others, enhancing their accuracy and utility. This holistic approach to luminosity measurement significantly improves the robustness and reliability of the data collected. In the central region, the Muon Drift Tubes (DT) luminometer is used covering a $\eta$ range from 0 to 1.2. The Pixel Cluster Counting (PCC) luminometer also operates in this region and extends into the forward region covering a range from 0 to 2.5. In the forward region, several devices are utilized. The Pixel Luminosity Telescope (PLT) covers the pseudorapidity range from 2.8 to 4.7. The Fast Beam Condition Monitor (BCM1F) functions in a range from 3.1 to 4.7. The Radiation Monitoring System for Environment and Safety (RAMSES) are located directly behind HF. The Hadron forward (HF) luminometer caters to the most forward region of the CMS detector \cite{Sirunyan:2759951}, functioning in the pseudorapidity range from 2.9 to 5.2 as shown in Fig.~\ref{fig:lumino_cms_pseudo}.

@@

\item Pixel Luminosity Telescope (PLT) : PLT is positioned 1.75 meters away from the interaction point, the PLT is an ensemble of 16 "telescopes." These are symmetrically distributed around the beam pipe at both extremities of the detector, with each telescope being composed of three individual silicon sensor planes. A defining feature of the PLT's functionality is the measurement of per-bunch instantaneous luminosity. This is accomplished by enumerating events in which all three sensor planes in a given telescope simultaneously register a hit. A specialized readout mechanism operating at the LHC's full bunch-crossing frequency of 40 MHz is employed to carry out this task. The PLT has provisions for reading out full pixel data, albeit at lower frequencies. This less frequent, yet detailed data readout serves multiple purposes. It is instrumental in determining calibrations, implementing corrections, and evaluating systematic uncertainties associated with both online and offline measurements. The PLT operates in what is called the "fast-OR" mode, which detects charged particles produced in a single bunch crossing of the proton beams. In this mode, the PLT triggers on any pixel in the detector that registers a signal above a certain threshold. This signal is called the "fast-OR signal" and indicates that a charged particle has passed through the detector. The PLT uses the information from all three detector planes to reconstruct the trajectories of the charged particles produced in the collisions. This is done by analyzing the signals from the individual pixels and using algorithms to determine the path of the particle through the detector.
  %The PLT consists of four layers of silicon pixel sensors arranged in a stack, with each layer containing over 8 million individual pixels. The first three layers are located very close to the interaction point where the proton beams collide, and the fourth layer is located further away. The innermost layer is located 2.9 cm from the interaction point. The second layer is located 3.9 cm from the interaction point. The third layer is located 4.9 cm from the interaction point. The outermost layer is located 10.2 cm from the interaction point \cite{Ayala:2861814}. The pixel sensors are designed to detect charged particles produced in the collisions as they pass through the detector layers. The PLT operates in what is called the "fast-OR" mode, which detects charged particles produced in a single bunch crossing of the proton beams. In this mode, the PLT triggers on any pixel in the detector that registers a signal above a certain threshold. This signal is called the "fast-OR signal" and indicates that a charged particle has passed through the detector. The PLT uses the information from all four detector planes to reconstruct the trajectories of the charged particles produced in the collisions. This is done by analyzing the signals from the individual pixels and using algorithms to determine the path of the particle through the detector.
  %To measure the luminosity of the collisions, the PLT counts the number of charged particles produced in each bunch crossing, and uses this information to calculate the rate of collisions. The luminosity is then calculated by dividing the collision rate by the cross-section of the proton-proton interactions.
  The PLT uses a specialized algorithm called the "triple coincidence" to identify charged particles that are produced in the collisions and to reject background noise. It requires a charged particle to be detected in all three planes, which corresponds to a "triple coincidence" detection. The rate of triple coincidences is proportional to the luminosity. This helps to reduce the contribution of background noise and improve the precision of the luminosity measurement \cite{Lujan:2017kvh}.

\item Fast Beam Conditions Monitor (BCM1F) : The BCM1F is specifically designed to detect the ionization produced by charged particles passing through its diamond sensors, and to convert this signal into a measurement of the beam intensity and luminosity. During Run 2, the BCM1F consisted of 24 diamond sensors, each with an active area of 1 cm², arranged in six rings around the beam pipe. The diamond sensors are positioned very close to the CMS detector, just 10 cm away from the beam pipe. This allows the sensors to detect the charged particles produced by the proton-proton collisions. When charged particles pass through the diamond sensors, they ionize the carbon atoms in the diamond crystal, producing electrons and holes. The electrons and holes are then collected by metal electrodes embedded in the diamond, producing a voltage signal proportional to the ionization produced by the charged particles. The voltage signal is then amplified by fast electronics and converted into a digital signal, which is sent to the CMS data acquisition system for further analysis. To measure the luminosity, the BCM1F counts the number of charged particles passing through each diamond sensor per unit time. The signal from each sensor is integrated over a short time interval (usually 25 ns), corresponding to the time it takes for a single proton bunch to pass through the sensors. The integrated signal is then converted into a measure of the instantaneous luminosity, which is proportional to the product of the number of charged particles per bunch and the collision frequency.

@@

\item Hadron Forward Calorimeter: The Hadron Forward (HF) Calorimeter is a critical component of the CMS experiment. It is a subdetector that measures the energy, position and arrival time of particles produced during proton-proton collisions. The HF Calorimeter is designed specifically to measure hadron particles, which are particles made up of quarks and gluons, in the forward region (i.e., close to the beamline) of the CMS detector. The HF Calorimeter is located at the very front and back of the CMS detector, at a distance of about 11.2 meters from the interaction point. This placement allows it to cover a pseudorapidity range of 2.9 to 5.2, which is important for measuring particles produced at small angles relative to the beamline. %The HF Calorimeter is a sampling calorimeter, which means it consists of alternating layers of active and passive materials.
The active material, in this case, is scintillating quartz fibers, which emit light when they interact with charged particles. The passive material is steel absorber plates, which serve to initiate hadron showers, causing the incident hadron to interact and create secondary particles. As these secondary particles pass through the scintillating fibers, they produce light, which is collected by photodetectors and converted into an electrical signal proportional to the particle's energy. Two separate algorithm are used for luminosity measurement in the HF Calorimeter, HFOC and HFET based on the Zero Counting (ZC) method and the sum of transverse energy methods respectively. %\cite{CMS-PAS-LUM-15-001}.

%The Zero Counting method is a threshold-based approach for estimating the luminosity. In this method, the total number of inelastic collisions, which produce new particles, is counted by measuring the energy deposited in the HF Calorimeter. To separate inelastic collisions from background noise and other processes, a threshold on the energy deposition is applied. During data-taking, the HF Calorimeter measures the energy deposited in each bunch crossing (when two bunches of protons pass through each other in the LHC). When the energy deposition in both HF Calorimeter sides (HF+ and HF-) is below a certain threshold, it is considered a "zero count" event, which means no inelastic collision has occurred. The number of zero count events is subtracted from the total number of bunch crossings, giving the number of inelastic collisions. The luminosity can then be calculated by dividing the number of inelastic collisions by the integrated time and the inelastic cross-section. The ZC method has the advantage of being robust against pile-up (simultaneous multiple proton-proton collisions) since it relies on counting the events where no energy is deposited. However, it may be affected by detector noise and other background processes that can lead to a non-zero energy deposition.
The zero counting method is used by particle colliders to measure the luminosity. In this method, the luminosity is estimated by counting the number of events with zero particles detected in a specific region of the detector. The basic idea behind the zero counting method is that the probability of having zero particles detected in a given region of the detector is proportional to the Poisson probability of having no collisions in that region. This probability can be related to the luminosity of the collisions through the following equation:

\begin{equation}
L = \frac{-ln(N/N_0)}{\sigma_{vis}}
\end{equation}

where L is the luminosity, N is the number of events with zero particles detected in the region of interest, $N_0$ is the total number of events, $\sigma_{vis}$ is the visible cross-section. One advantage is that it is a simple and robust method that does not require sophisticated detectors or calibration procedures. However, it can be affected by background events, such as cosmic rays or beam-gas interactions. \\
The HFET algorithm measures luminosity by summing transverse energy deposits in the HF calorimeter cells. It relies on the direct proportionality between the total transverse energy and the particle flux from the interaction point. Accurate calibration of calorimeter cells is crucial for this method, as it ensures that the aggregate transverse energy accurately reflects the luminosity of the particle collisions \cite{Sirunyan:2759951.

@@

\item Muon drift tubes: The DT system is part of the CMS muon system, %which also includes the Resistive Plate Chambers (RPCs) and the Cathode Strip Chambers (CSCs).
  Drift Tubes (DTs) in the CMS experiment are designed to accurately measure the trajectory and momentum of muons produced in proton-proton collisions at the LHC.
  %The DT system is located in the barrel region of the CMS detector, covering the pseudorapidity range of $|\eta| < 1.2$.
  The DT system consist of rectangular drift cells filled with a gas mixture, typically of argon and carbon dioxide. The use of DT muon rates for luminosity measurement is based on the idea that the rate of single muon production is proportional to the instantaneous luminosity. Single muons comes from decay of inclusive, Z, W, top and jets.
%The DT system in the CMS detector is used for muon detection and precise position measurement. Drift Tubes are essentially gas-filled detectors. When muons pass through these tubes, they ionize the gas. The resulting ions and electrons drift under the influence of an electric field to anode wires located in the center of the tubes, producing a signal. By measuring the drift time, one can infer the position of the muon's trajectory.
  %To use the DT muon rates as a luminometer, one would need to count the rate of detected muons as a function of the delivered luminosity.
However, the DT muon rate luminometer is not independently calibrated during the van der Meer (vdM) scan due to low measured rate. It is cross calibrated with PCC or HF.

@@

\begin{figure}[!htp]
\centering
\includegraphics[width=0.9\textwidth]{ashish_thesis/luminometer_cms_pseudo.png}
\caption[CMS luminometers Location]{%
    Cross section of the CMS detector in the r-z plane showing all Run 2 luminometer PLT, BCM1F, DTs, and HF location. The two RAMSES monitors used as a luminometer in Run 2 are located directly behind HF \cite{Sirunyan:2759951}. 
}
\label{fig:lumino_cms_pseudo}
\end{figure}

@@

The pixel detector is one of the subdetectors of the CMS experiment that is closest to the interaction point where the proton beams collide. It consists of four barrel layers (L1-L4) at radii of 29, 68, 109 and 160 mm and three forward disks (D1-D3) on each end at distances of 291, 396, and 516 mm of pixel sensors from the center of the detector as shown in Fig.~\ref{fig:phaseI_upgrade}. It is built from 1856 silicon sensor modules, 1184 modules in the barrel layers (BPIX) and 672 modules for the forward disks (FPIX) \cite{Adam:2748381}. Modules for barrel layers and forward disks are shown in Fig.~\ref{fig:pixelmodule}. Table~\ref{tab:modpix} shows the number of modules in each pixel layer and disk. The modules in BPIX L1 are designed to handle high particle densities and radiation levels. Each module in this layer consists of a matrix of 4160 pixels, arranged in 160 rows and 26 columns. The pixel size in BPIX L1 is approximately 100x150 µm², which allows for high-resolution position measurements of charged particles. The modules in BPIX L2-4 layers are designed to withstand lower radiation levels compared to BPIX L1. Each module in BPIX L2-4 consists of a matrix of 66560 pixels, arranged in 260 rows and 256 columns. The pixel size in BPIX L2-4 is also approximately 100x150 µm². The FPIX modules are also designed to handle high particle densities and radiation levels. Each module in the FPIX consists of a matrix of 16128 pixels, arranged in 160 rows and 101 columns with same pixel size as layer pixels.
%It provide high-precision position measurements of the charged particles produced in the collisions. These position measurements are used to reconstruct the trajectories of the particles and to identify the particles themselves.
The pixel detector has a high position resolution, with a spatial resolution of 4.8  microns in the transverse direction and 7.99 microns in the longitudinal direction. The pixel detector also has a high time resolution, with a readout time of 25 nanoseconds. The pixel detector is capable of detecting charged particles with a wide range of energies, from a few hundred MeV to several TeV. The pixel detector is able to operate at very high particle fluxes, with a hit rate of up to $100 MHz/cm^2$. To achieve this high rate capability, the pixel detector uses fast readout electronics.
%which can read out the signals from the pixels in a few microseconds.
The pixel detector is also designed to operate at a high radiation level, with a dose of up to $5 \times 10^{14} neutrons/cm^2$ expected over the lifetime of the detector. To mitigate the effects of radiation damage, the pixel detector uses sensors made of high-purity silicon and a cooling system that keeps the detector at a temperature of -20°C. The pixel detector is  used as a primary luminometer. The luminosity measurement is based on counting the number of pixel clusters produced in the detector by the proton-proton collisions.  However, there are several challenges to accurately measure the luminosity with the pixel detector. One challange is the multiple proton-proton collisions per bunch crossing (pileup). Another challenge is the possible non-linearity or instabilities of the pixel detector response, which can lead to inaccuracies in the luminosity measurement. Therefore, the motivation of this research work is to develop and test more precise techniques for measuring the luminosity with the pixel detector.

@@

\section{Pixel cluster counting (PCC) method}

The Pixel Cluster Counting (PCC) method is one of the main luminosity measurement technique used at the CMS experiment. The PCC method is based on counting the number of pixel clusters produced by the collisions in the CMS pixel detector, which is a high-resolution silicon detector located close to the collision point. The pixel detector is made up of more than 124 million pixels. The detector is designed to detect charged particles produced in the collisions, and the pixel clusters are formed when several neighboring pixels are hit by a single particle. The PCC method is based on the assumption that the number of pixel clusters produced is proportional to the number of collisions, and therefore to the luminosity. The proportionality constant is called the pixel detector visible cross section, and it can be determined with a calibration method described in section 3.1. The PCC method exclusively utilizes offline luminosity measurement, comprising two stages: the "prompt-lumi," which is delayed by a few days and employs a non-optimized module list, and the "offline reprocessed luminosity" that uses an optimized module list for enhanced accuracy. During Run 2, the PCC online role was confined to recording pixel information, with cluster reconstruction occurring offline. However, in Run 3, this process evolved, allowing clusters to be reconstructed online within the High-Level Trigger (HLT) system, subsequently recorded for detailed offline analysis.
%The PCC method is implemented in two stages: the online and offline stages. In the online stage, the pixel clusters are counted in real time by the CMS data acquisition system, which uses a custom-designed ASIC chip to process the pixel detector signals. %The online luminosity measurement is therefore available within seconds of the collisions, and it is used to monitor the stability of the detector and to provide feedback to the LHC operators.
%In the offline stage, the raw data from the pixel detector is analyzed using dedicated software to determine the pixel clusters and to calculate the luminosity. The offline luminosity measurement is more accurate than the online measurement, as it takes into account various corrections and calibrations that are not possible in real time. %The offline luminosity measurement is also used to determine the absolute luminosity scale, which is obtained from beam parameters.
There are several steps involved in the PCC method for measuring luminosity like data acquisition, pixel cluster reconstruction, pixel cluster counting as described below.

@@

\item The number of pixel clusters is counted over a fixed time intervals, known as lumi section. One lumi section is 23.36 seconds time interval. In the pixel cluster counting process, events of interest are selected from the large dataset obtained from the CMS experiment. This is typically done using zero bias triggers that randomly select events that are coming from filled bunch crossing.
  %Once the events of interest have been selected, reconstruction the pixel clusters in each event is done using the steps described in the previous bullet point.
  %This involves identifying individual pixel hits, identifying seed pixels, building pixel clusters around each seed pixel, and parameterizing the clusters.
  After the pixel clusters have been reconstructed in each event, the next step is to count the number of clusters in each event. This can be done using simple counting algorithms that count the number of clusters above a certain energy threshold. %or more sophisticated algorithms that use machine learning techniques to identify and count clusters.

@@

\item The number of pixel clusters is used to calculate the luminosity using the following formula:

\begin{equation}
 L_b = \frac{N_{cl} f_{rev} } {\sigma_{vis}}   
\end{equation}

where $L_b$ is the bunch luminosity, $N_{cl}$ is the number of clusters per bunch crossing, $f_{rev}$ is the revolution frequency of the LHC, and $\sigma_{vis}$ is the visible cross section of the pixel detector obtained from calibration.

>>>>

\chapter{Luminosity measurement at the CMS experiment}  %Title of the First Chapter

@@

The CMS experiment employs several devices to measure luminosity, each covering specific pseudorapidity ranges and regions within the CMS detector \cite{pas_18}. The use of several luminometers as shown in Fig.~\ref{fig:lumino_cms} offers a multitude of advantages. Primarily, it ensures redundancy and acts as a backup for potential device failures, ensuring that data collection can proceed without interruption. Additionally, having multiple luminometers allows for cross-checking, providing greater consistency in results and allowing for more reliable data interpretation. These different devices enable better control over systematic uncertainties, as any discrepancies can be identified and adjusted for. Moreover, each device employs different method to measure luminosity, providing complementary information that enriches the overall understanding of particle interactions. Some luminometers recording low rates can be calibrated using others, enhancing their accuracy and utility. This holistic approach to luminosity measurement significantly improves the robustness and reliability of the data collected. In the central region, the Muon Drift Tubes (DT) luminometer is used covering a $|\eta|$ range from 0 to 1.2. The Pixel Cluster Counting (PCC) luminometer also operates in this region and extends into the forward region covering a range from 0 to 2.5. In the forward region, several devices are utilized. The Pixel Luminosity Telescope (PLT) covers the pseudorapidity range from 2.8 to 4.7. The Fast Beam Condition Monitor (BCM1F) functions in a range from 3.1 to 4.7. The Hadron forward (HF) luminometer caters to the most forward region of the CMS detector \cite{Sirunyan:2759951}, functioning in the pseudorapidity range from 2.9 to 5.2 as shown in Fig.~\ref{fig:lumino_cms_pseudo}. The Radiation Monitoring System for Environment and Safety (RAMSES) are located directly behind HF.

@@

\item Pixel Luminosity Telescope (PLT) : PLT is positioned 1.75 meters away from the interaction point. The PLT is an ensemble of 16 "telescopes". These are symmetrically distributed around the beam pipe at both sides of the detector, with each telescope being composed of three individual silicon sensor planes. A defining feature of the PLT functionality is the measurement of per-bunch instantaneous luminosity. This is accomplished by enumerating events in which all three sensor planes in a given telescope simultaneously register a hit. A specialized readout mechanism operating at the LHC full bunch-crossing frequency of 40 MHz is employed to carry out this task. The PLT has provisions for reading out full pixel data, albeit at lower frequencies. This less frequent, yet detailed data readout serves multiple purposes. It is instrumental in determining calibrations, implementing corrections, and evaluating systematic uncertainties associated with both online and offline measurements. The PLT operates in what is called the "fast-OR" mode, which detects charged particles produced in a single bunch crossing of the proton beams. In this mode, the PLT triggers on any pixel in the detector that registers a signal above a certain threshold. This signal is called the "fast-OR signal" and indicates that a charged particle has passed through the detector. The PLT uses the information from all three detector planes to reconstruct the trajectories of the charged particles produced in the collisions. This is done by analyzing the signals from the individual pixels and using algorithms to determine the path of the particle through the detector.
  %The PLT consists of four layers of silicon pixel sensors arranged in a stack, with each layer containing over 8 million individual pixels. The first three layers are located very close to the interaction point where the proton beams collide, and the fourth layer is located further away. The innermost layer is located 2.9 cm from the interaction point. The second layer is located 3.9 cm from the interaction point. The third layer is located 4.9 cm from the interaction point. The outermost layer is located 10.2 cm from the interaction point \cite{Ayala:2861814}. The pixel sensors are designed to detect charged particles produced in the collisions as they pass through the detector layers. The PLT operates in what is called the "fast-OR" mode, which detects charged particles produced in a single bunch crossing of the proton beams. In this mode, the PLT triggers on any pixel in the detector that registers a signal above a certain threshold. This signal is called the "fast-OR signal" and indicates that a charged particle has passed through the detector. The PLT uses the information from all four detector planes to reconstruct the trajectories of the charged particles produced in the collisions. This is done by analyzing the signals from the individual pixels and using algorithms to determine the path of the particle through the detector.
  %To measure the luminosity of the collisions, the PLT counts the number of charged particles produced in each bunch crossing, and uses this information to calculate the rate of collisions. The luminosity is then calculated by dividing the collision rate by the cross-section of the proton-proton interactions.
  The PLT uses a specialized algorithm called the "triple coincidence" to identify charged particles that are produced in the collisions and to reject background noise. It requires a charged particle to be detected in all three planes, which corresponds to a "triple coincidence" detection. The rate of triple coincidences is proportional to the luminosity. This helps to reduce the contribution of background noise and improve the precision of the luminosity measurement \cite{Lujan:2017kvh}.

\item Fast Beam Conditions Monitor (BCM1F) : The BCM1F is specifically designed to detect the ionization produced by charged particles passing through its diamond sensors, and to convert this signal into a measurement of the beam intensity and luminosity. During Run 2, the BCM1F consisted of 24 diamond sensors, each with an active area of 1 cm², arranged in six rings around the beam pipe. The diamond sensors are positioned very close to the CMS detector, just 10 cm away from the beam pipe. This allows the sensors to detect the charged particles produced by the proton-proton collisions. When charged particles pass through the diamond sensors, they ionize the carbon atoms in the diamond crystal, producing electrons and holes. The electrons and holes are then collected by metal electrodes embedded in the diamond, producing a voltage signal proportional to the ionization produced by the charged particles. The voltage signal is then amplified by fast electronics and converted into a digital signal, which is sent to the CMS data acquisition system for further analysis. To measure the luminosity, the BCM1F counts the number of charged particles passing through each diamond sensor per unit time. The signal from each sensor is integrated over a short time interval (usually 25 ns), corresponding to the time it takes for a single proton bunch to pass through the sensors. The integrated signal is then converted into a measure of the instantaneous luminosity, which is proportional to the product of the number of charged particles per bunch and the collision frequency. Run3 FBCM employs silicon-pad sensors located inside the cold volume of the Tracker, leveraging the zero-counting algorithm for luminosity measurement and also offering timing information with a few nanoseconds precision for beam-induced background measurement.

@@

\item Hadron Forward Calorimeter: The Hadron Forward (HF) Calorimeter is a critical component of the CMS experiment. It is a subdetector that measures the energy, position and arrival time of particles produced during proton-proton collisions. The HF Calorimeter is designed specifically to measure hadron particles, which are particles made up of quarks and gluons, in the forward region (i.e., close to the beamline) of the CMS detector. The HF Calorimeter is located at the very front and back of the CMS detector, at a distance of about 11.2 meters from the interaction point. This placement allows it to cover a pseudorapidity range of 2.9 to 5.2, which is important for measuring particles produced at small angles relative to the beamline. %The HF Calorimeter is a sampling calorimeter, which means it consists of alternating layers of active and passive materials.
The active material, in this case, is scintillating quartz fibers, which emit light when they interact with charged particles. The passive material is steel absorber plates, which serve to initiate hadron showers, causing the incident hadron to interact and create secondary particles. As these secondary particles pass through the scintillating fibers, they produce light, which is collected by photodetectors and converted into an electrical signal proportional to the particle's energy. Two separate algorithms are used for luminosity measurement in the HF Calorimeter, HFOC and HFET based on the Zero Counting (ZC) method and the sum of transverse energy methods respectively. %\cite{CMS-PAS-LUM-15-001}.

%The Zero Counting method is a threshold-based approach for estimating the luminosity. In this method, the total number of inelastic collisions, which produce new particles, is counted by measuring the energy deposited in the HF Calorimeter. To separate inelastic collisions from background noise and other processes, a threshold on the energy deposition is applied. During data-taking, the HF Calorimeter measures the energy deposited in each bunch crossing (when two bunches of protons pass through each other in the LHC). When the energy deposition in both HF Calorimeter sides (HF+ and HF-) is below a certain threshold, it is considered a "zero count" event, which means no inelastic collision has occurred. The number of zero count events is subtracted from the total number of bunch crossings, giving the number of inelastic collisions. The luminosity can then be calculated by dividing the number of inelastic collisions by the integrated time and the inelastic cross-section. The ZC method has the advantage of being robust against pile-up (simultaneous multiple proton-proton collisions) since it relies on counting the events where no energy is deposited. However, it may be affected by detector noise and other background processes that can lead to a non-zero energy deposition.
The zero counting method is used by particle colliders to measure the luminosity. In this method, the luminosity is estimated by counting the number of events with zero particles detected in a specific region of the detector. The basic idea behind the zero counting method is that the probability of having zero particles detected in a given region of the detector is proportional to the Poisson probability of having no collisions in that region. This probability can be related to the instantaneous luminosity of the collisions through the following equations,

We consider the Poisson distribution, which gives the probability of observing a certain number of events, $n$, when the average number of events is $\lambda$:

\[
P(n; \lambda) = \frac{e^{-\lambda} \lambda^n}{n!}
\]

For zero-counting, we are interested in the probability of observing no events ($n=0$). Substituting $n=0$ into the Poisson distribution equation, we get:

\[
P(0; \lambda) = \frac{e^{-\lambda} \lambda^0}{0!} = e^{-\lambda}
\]

The probability of observing zero events, $P(0; \lambda)$, is related to the average number of events $\lambda$, which can be determined by taking the negative natural logarithm of $P(0; \lambda)$:

\[
\lambda = -\ln(P(0; \lambda))
\]

The average number of events per bunch crossing, $\lambda$, is related to the instantaneous luminosity, $L_{inst}$, and the total visible cross-section, $\sigma_{\text{vis}}$, by:

\[
\lambda = L_{inst} \cdot \sigma_{\text{vis}}
\]

Substituting $\lambda$ in terms of $L$ and $\sigma_{\text{vis}}$ into the equation involving $P(0; \lambda)$, we get:

\[
L_{inst} \cdot \sigma_{\text{vis}} = -\ln(P(0; \lambda))
\]

Solving for $L_{inst}$ gives us the equation for instantaneous luminosity based on the zero-counting method:

\[
L_{inst} = \frac{-\ln(P(0; \lambda))}{\sigma_{\text{vis}}}
\]

The probability $P(0; \lambda)$ can be written as the ratio $\frac{N}{N_0}$ because it is the fraction of the total number of events where zero particles are detected:
\[
P(0; \lambda) = \frac{N}{N_0}
\]

Substituting $P(0; \lambda) = \frac{N}{N_0}$ into the equation for $L_{inst}$ gives us:

\begin{equation}
L_{inst} = \frac{-ln(N/N_0)}{\sigma_{vis}}
\end{equation}

where $L_{inst}$ is the luminosity, N is the number of events with zero particles detected in the region of interest, $N_0$ is the total number of events, $\sigma_{vis}$ is the visible cross-section. One advantage is that it is a simple and robust method that does not require sophisticated detectors or calibration procedures. However, it can be affected by background events, such as cosmic rays or beam-gas interactions. \\
The HFET algorithm measures luminosity by summing transverse energy deposits in the HF calorimeter cells. It relies on the direct proportionality between the total transverse energy and the particle flux from the interaction point. Accurate calibration of calorimeter cells is crucial for this method, as it ensures that the aggregate transverse energy accurately reflects the luminosity of the particle collisions \cite{Sirunyan:2759951.

@@

\item Muon drift tubes: The DT system is part of the CMS muon system. %which also includes the Resistive Plate Chambers (RPCs) and the Cathode Strip Chambers (CSCs).
  DTs in the CMS experiment are designed to accurately measure the trajectory and momentum of muons produced in proton-proton collisions at the LHC.
  %The DT system is located in the barrel region of the CMS detector, covering the pseudorapidity range of $|\eta| < 1.2$.
  The DT system consists of rectangular drift cells filled with a gas mixture, typically of argon and carbon dioxide. The use of DT muon rates for luminosity measurement is based on the idea that the rate of muon production is proportional to the instantaneous luminosity. Muons come from the decay of inclusive, Z, W, top and jets.
%The DT system in the CMS detector is used for muon detection and precise position measurement. Drift Tubes are essentially gas-filled detectors. When muons pass through these tubes, they ionize the gas. The resulting ions and electrons drift under the influence of an electric field to anode wires located in the center of the tubes, producing a signal. By measuring the drift time, one can infer the position of the muon's trajectory.
  %To use the DT muon rates as a luminometer, one would need to count the rate of detected muons as a function of the delivered luminosity.
However, the DT muon rate luminometer is not independently calibrated during the van der Meer (vdM) scan due to low measured rate. It is cross calibrated with PCC or HF.

@@

\begin{figure}[!htp]
\centering
\includegraphics[width=0.9\textwidth]{ashish_thesis/luminometer_cms_pseudo.png}
\caption[CMS luminometers Location]{%
    Cross section of the CMS detector in the r-z plane showing all Run 2 luminometers PLT, BCM1F, DTs, and HF location. The two RAMSES monitors used as a luminometer in Run 2 are located directly behind HF \cite{Sirunyan:2759951}. 
}
\label{fig:lumino_cms_pseudo}
\end{figure}

@@

The pixel detector is one of the subdetectors of the CMS experiment that is closest to the interaction point where the proton beams collide. It consists of four barrel layers (L1-L4) at radii of 29, 68, 109 and 160 mm and three forward disks (D1-D3) on each end at distances of 291, 396, and 516 mm from the center of the detector as shown in Fig.~\ref{fig:phaseI_upgrade}. It is built from 1856 silicon sensor modules, 1184 modules in the barrel layers (BPIX) and 672 modules in the forward disks (FPIX) \cite{Adam:2748381}. Modules for barrel layers and forward disks are shown in Fig.~\ref{fig:pixelmodule}. Table~\ref{tab:modpix} shows the number of modules in each pixel layer and disk. The modules in BPIX L1 are designed to handle high particle densities and radiation levels. Each module in this layer consists of a matrix of 4160 pixels, arranged in 160 rows and 26 columns. The pixel size in BPIX L1 is approximately 100x150 µm², which allows for high-resolution position measurements of charged particles. The modules in BPIX L2-4 layers are designed to withstand lower radiation levels compared to BPIX L1. Each module in BPIX L2-4 consists of a matrix of 66560 pixels, arranged in 260 rows and 256 columns. The pixel size in BPIX L2-4 is also approximately 100x150 µm². The FPIX modules are also designed to handle high particle densities and radiation levels. Each module in the FPIX consists of a matrix of 16128 pixels, arranged in 160 rows and 101 columns with same pixel size as barrel layer pixels.
%It provide high-precision position measurements of the charged particles produced in the collisions. These position measurements are used to reconstruct the trajectories of the particles and to identify the particles themselves.
The pixel detector has a high position resolution, with a spatial resolution of 4.8  microns in the transverse direction and 7.99 microns in the longitudinal direction. The pixel detector also has a high time resolution, with a readout time of 25 nanoseconds. The pixel detector is capable of detecting charged particles with a wide range of energies, from a few hundred MeV to several TeV. The pixel detector is able to operate at very high particle fluxes, with a hit rate of up to $100 \: MHz/cm^2$. To achieve this high rate capability, the pixel detector uses fast readout electronics.
%which can read out the signals from the pixels in a few microseconds.
The pixel detector is also designed to operate at a high radiation level, with a dose of up to $5 \times 10^{14} neutrons/cm^2$ expected over the lifetime of the detector. To mitigate the effects of radiation damage, the pixel detector uses sensors made of high-purity silicon and a cooling system that keeps the detector at a temperature of -20°C. The pixel detector is  used as a primary luminometer. The luminosity measurement is based on counting the number of pixel clusters produced in the detector by the proton-proton collisions.  However, there are several challenges to accurately measure the luminosity with the pixel detector. One challange is the multiple proton-proton collisions per bunch crossing (pileup). Another challenge is the possible non-linearity or instabilities of the pixel detector response, which can lead to inaccuracies in the luminosity measurement. Therefore, the motivation of this research work is to develop and test more precise techniques for measuring the luminosity with the pixel detector.

@@

\section{Pixel cluster counting method}

The Pixel Cluster Counting (PCC) method is one of the main luminosity measurement techniques used at the CMS experiment. The PCC method is based on counting the number of pixel clusters produced by the collisions in the CMS pixel detector, which is a high-resolution silicon detector located close to the collision point. The pixel detector is made up of more than 124 million pixels. The detector is designed to detect charged particles produced in the collisions, and the pixel clusters are formed when several neighboring pixels are hit by a single particle. The PCC method is based on the assumption that the number of pixel clusters produced is proportional to the number of collisions, and therefore to the luminosity. The proportionality constant is called the pixel detector visible cross section, and it can be determined with a calibration method described in section 3.1. The PCC method exclusively utilizes offline luminosity measurement, comprising two stages: the "prompt-lumi," which is delayed by a few days and employs a non-optimized module list, and the "offline reprocessed luminosity" that uses an optimized module list for enhanced accuracy. During Run 2, the PCC online role was confined to recording pixel information, with cluster reconstruction occurring offline. However, in Run 3, this process evolved, allowing clusters to be reconstructed online within the High-Level Trigger (HLT) system, subsequently recorded for detailed offline analysis.
%The PCC method is implemented in two stages: the online and offline stages. In the online stage, the pixel clusters are counted in real time by the CMS data acquisition system, which uses a custom-designed ASIC chip to process the pixel detector signals. %The online luminosity measurement is therefore available within seconds of the collisions, and it is used to monitor the stability of the detector and to provide feedback to the LHC operators.
%In the offline stage, the raw data from the pixel detector is analyzed using dedicated software to determine the pixel clusters and to calculate the luminosity. The offline luminosity measurement is more accurate than the online measurement, as it takes into account various corrections and calibrations that are not possible in real time. %The offline luminosity measurement is also used to determine the absolute luminosity scale, which is obtained from beam parameters.
There are several steps involved in the PCC method for measuring luminosity like data acquisition, pixel cluster reconstruction and pixel cluster counting as described below.

@@

\item The number of pixel clusters is counted over a fixed time interval, known as lumi section. One lumi section is 23.36 seconds time interval. In the pixel cluster counting process, events of interest are selected from the large dataset obtained from the CMS experiment. This is typically done using zero bias triggers that randomly select events that are coming from filled bunch crossings.
  %Once the events of interest have been selected, reconstruction the pixel clusters in each event is done using the steps described in the previous bullet point.
  %This involves identifying individual pixel hits, identifying seed pixels, building pixel clusters around each seed pixel, and parameterizing the clusters.
  After the pixel clusters have been reconstructed in each event, the next step is to count the number of clusters in each event. This can be done using simple counting algorithms that count the number of clusters above a certain energy threshold. %or more sophisticated algorithms that use machine learning techniques to identify and count clusters.

@@

\item The number of pixel clusters is used to calculate the instantaneous luminosity using the following formula:

\begin{equation}
 L_{inst} = \frac{N_{cl} f_{rev} } {\sigma_{vis}}   
\end{equation}

where $L_{inst}$ is the bunch luminosity, $N_{cl}$ is the number of clusters per bunch crossing, $f_{rev}$ is the revolution frequency of the LHC, and $\sigma_{vis}$ is the visible cross section of the pixel detector obtained from calibration.
